Accelerating Optimal Power Flow with GPUs: SIMD Abstraction of Nonlinear Programs and Condensed-Space Interior-Point Methods

### 2023-07-27 paper/main.tex
finished section 3

\subsection{Numerical Optimization}\label{sec:numopt}
We consider NLPs of the following form:
\begin{subequations}\label{eqn:cpt}
  \begin{align}
    \min_{x^\flat \leq x \leq x^\sharp}\;
    & f(x)\\
    \st\;
    & g(x) =0. \label{eqn:cpt-con}
  \end{align} 
\end{subequations}
Numerous solution algorithms have been developed in the NLP
literature. In terms of strategies to deal with inequality
constraints, the NLP solution algorithms can be broadly classified
into active-set methods and IPMs \cite{nocedal2006numerical}.
Active-setmethods aim to find the set of active consraints associated
with the optimal solution in a combinatorial manner, while IPMs
replace inequality constraints with smooth barrier functions, enabling
the use of Newton-type second-order algorithms. IPMs are known to be
more scalable for problems with a large number of constraints and
suitable for parallelization, thanks to the fixed sparsity pattern of
the KKT matrix. Given these advantages, we have chosen IPMs as the
backbone algorithm for developing our optimization methods on GPUs.

@@

We highlight that the following, arguably quite common, computational
patterns are particularly effective when implemented on GPUs:
\begin{align}
  y&\leftarrow \left[\phi(x; q_j)\right]_{j\in [J]}\tag{Pattern 1}\label{eqn:pattern-1}\\
  o&\leftarrow  \mathop{\rm{Op}}_{i\in [I]} \psi(x;p_i) \tag{Pattern 2}\label{eqn:pattern-2}\\
  x&\leftarrow  \chi_{s_1}\circ\cdots\circ \chi_{s_K}(x) \tag{Pattern 3}\label{eqn:pattern-3}
\end{align}
Here, $\psi:\mathbb{R}^{n_x}\times \mathbb{R}^{n_{p}}\rightarrow
\mathbb{R}$ , $\phi:\mathbb{R}^{n_x}\times
\mathbb{R}^{n_{q}}\rightarrow \mathbb{R}$, and
$\chi:\mathbb{R}^{n_x}\times \mathbb{R}^{n_{s}}\rightarrow
\mathbb{R}^{n_x}$ are simple instructions that require only a small number
of operations; $\mathop{\rm{Op}}$ is a monoid operator on
$\mathbb{R}\cup\{+\infty,-\infty\}$, such as addition, multiplication,
maximum, and minimum; and the notation $\chi_{s_k}(x):=\chi(x,s_k)$ is used, and
assume that $\circ$ is commutative on $\{\chi_{s_k}(\cdot)\}_{\forall
s_k}$. \ref{eqn:pattern-1} is typically most effective on GPUs, where
each thread employed can operate independently without needing to
simultaneously manipulate the same device memory location. While
\ref{eqn:pattern-2} and \ref{eqn:pattern-3} is less effective as they
require such a manipulation, but they still can be significantly
faster than operations in CPUs, as substantial part of the operation
can still be parallelized by the use of buffers. In simple cases, the
implementation of these operations can be performed with the standard
{\tt map} and {\tt mapreduce} programming models. However, in more
complex cases, especially for \ref{eqn:pattern-3}, the
implementation may require pre-inspection of memory write-access
patterns and the use of custom kernels.

@@

Implementing kernel functions for the above computational patterns in
the Julia Language is straightforward, as it provides excellent
high-level interfaces for array and kernel programming for GPU
arrays. The code can even be device-agnostic, thanks to the portable
programming capabilities brought by KernelAbstractions.jl. All
of the AD and optimization capabilities in our
tools, MadNLP.jl and SIMDiff.jl, are implemented in Julia Language,
by leveraging its kernel and array programming capabilities.

@@

\subsection{Abstraction}
The SIMD abstraction under consideration is as follows:
\begin{align}\label{eqn:prob}
  \min_{x^\flat\leq x \leq x^\sharp}
  & \sum_{l\in[L]}\sum_{i\in [I_l]} f^{(l)}(x; p^{(l)}_i)\\\nonumber
  \st\; &\forall m\in[M]:\\\nonumber
  &\begin{aligned}[t]
    \left[g^{(m)}(x; q_j)\right]_{j\in [J_m]} +\sum_{n\in [N_m]}\sum_{k\in [K_n]}h^{(n)}(x; s^{(n)}_{k}) =0,
  \end{aligned}
  \end{align}
where $f^{(\ell)}(\cdot,\cdot)$, $g^{(m)}(\cdot,\cdot)$, and
$h^{(n)}(\cdot,\cdot)$ are twice differentiable functions with respect
to the first argument, $\{\{p^{(k)}_i\}_{i\in [N_k]}\}_{k\in[K]}$,
$\{\{q^{(k)}_{i}\}_{i\in [M_l]}\}_{m\in[M]}$, and
$\{\{\{s^{(n)}_{k}\}_{k\in[K_n]}\}_{n\in[N_m]}\}_{m\in[M]}$ are
problem data, which can either be discrete or continuous.  We assume
that our functions $f^{(l)}(\cdot,\cdot)$, $g^{(m)}(\cdot,\cdot)$, and
$h^{(n)}(\cdot,\cdot)$ can be expressed in the form of computational
graphs of moderate length. One can observe that the problem in
\eqref{eqn:prob} is expressed by the computational patterns in Section
\ref{sec:gpu}. In particular, the objective function falls within \ref{eqn:pattern-1},
the first term in the constraint falls within \ref{eqn:pattern-2}, and the second term in the constraint falls within \ref{eqn:pattern-3}.
Accordingly, the evaluation and differentiation of the
model equations in \eqref{eqn:prob} is amenable to SIMD parallelism.

@@

\subsection{Parallel AD}
Many of the physics-based models, such as AC OPF, have highly repetitive
structure. One of the manifestations of it is that the mathematical
statement of the model is concise, even if the practical model may contain
millions of variables and constraints. This is possible due to the use of
repetition over a certain index and data sets. For example,
it suffices to use 15 computational patterns to fully specify the
AC OPF model. These patterns arise from (1) generation cost, (2) reference
bus voltage angle constraint, (3-6) active and reactive power flow (to
and from), (7) voltage angle difference constraint, (8-9) apparent
power flow limits (to and from), (10-11) power balance equations,
(12-13) generators' contributions to the power balance equations, and
(14-15) in/out flows contributions to the power balance
equations. However, such repetitive structure is not well exploited
during AD in the standard NLP modeling paradigms.  In
fact, without the SIMD abstraction, it is difficult for the AD package
to detect the parallelizable structure within the model, as it will
require the full inspection of the computational graph. By preserving
the repetitive structures in the model, one can facilitate the
implementation of parallelized derivative computations.

@@

SIMDiff.jl exploits the SIMD abstraction of the model equations to
save the computational cost spent for sparsity analysis.  This is
accomplished by applying sparsity analysis for the instruction for
each computational pattern and expand the obtained sparsity pattern
over the data over which the instruction is executed. Specifically,
the sparisty analysis code exploits Julia's multiple dispatch feature
to obtain parameterized sparsity pattern for each instruction, and the
obtained parameterized sparsity pattern is materialized once the data
array is given. This process drastically save the computational cost
for the sparsity analysis.

>>>>

\subsection{Numerical Optimization}\label{sec:numopt}
We consider NLPs of the following form:
\begin{subequations}\label{eqn:cpt}
  \begin{align}
    \min_{x^\flat \leq x \leq x^\sharp}\;
    & f(x)\\
    \st\;
    & g(x) =0. \label{eqn:cpt-con}
  \end{align} 
\end{subequations}
Numerous solution algorithms have been developed in the NLP
literature. In terms of strategies to deal with inequality
constraints, the NLP solution algorithms can be broadly classified
into active-set methods and IPMs \cite{nocedal2006numerical}.
Active-set methods aim to find the set of active constraints associated
with the optimal solution in a combinatorial manner, while IPMs
replace inequality constraints with smooth barrier functions, enabling
the use of Newton-type second-order algorithms. IPMs are known to be
more scalable for problems with a large number of constraints and
suitable for parallelization, thanks to the fixed sparsity pattern of
the KKT matrix. Given these advantages, we have chosen IPMs as the
backbone algorithm for developing our optimization methods on GPUs.

@@

We highlight that the following, arguably quite common, computational
patterns are particularly effective when implemented on GPUs:
\begin{align}
  y&\leftarrow \left[\phi(x; q_j)\right]_{j\in [J]}\tag{Pattern 1}\label{eqn:pattern-1}\\
  o&\leftarrow  \mathop{\rm{Op}}_{i\in [I]} \psi(x;p_i) \tag{Pattern 2}\label{eqn:pattern-2}\\
  x&\leftarrow  \chi_{s_1}\circ\cdots\circ \chi_{s_K}(x) \tag{Pattern 3}\label{eqn:pattern-3}
\end{align}
Here, $\psi:\mathbb{R}^{n_x}\times \mathbb{R}^{n_{p}}\rightarrow
\mathbb{R}$ , $\phi:\mathbb{R}^{n_x}\times
\mathbb{R}^{n_{q}}\rightarrow \mathbb{R}$, and
$\chi:\mathbb{R}^{n_x}\times \mathbb{R}^{n_{s}}\rightarrow
\mathbb{R}^{n_x}$ are simple instructions that require only a small number
of operations; $\mathop{\rm{Op}}$ is a monoid operator on
$\mathbb{R}\cup\{+\infty,-\infty\}$, such as addition, multiplication,
maximum, and minimum. In \ref{eqn:pattern-3}, we denote $\chi_{s_k}(x):=\chi(x,s_k)$ and
assume that $\circ$ is commutative for $\{\chi_{s_k}(\cdot)\}_{\forall
s_k}$. \ref{eqn:pattern-1} is typically most effective on GPUs, where
each thread employed can operate independently without needing to
simultaneously manipulate the same device memory location. While
\ref{eqn:pattern-2} and \ref{eqn:pattern-3} are less effective as they
require such manipulations, they still can be significantly
faster than operations in CPUs, as substantial part of the operation
can still be parallelized by the use of buffers. In simple cases, the
implementation of these operations can be performed with the standard
{\tt map} and {\tt mapreduce} programming models. However, in more
complex cases, especially for \ref{eqn:pattern-3}, the
implementation may require pre-inspection of memory write-access
patterns and the use of custom kernels.

@@

Implementing kernel functions for the above computational patterns in
the Julia Language is straightforward, as it provides excellent
high-level interfaces for array and kernel programming for GPU
arrays. The code can even be device-agnostic, thanks to the portable
programming capabilities brought by KernelAbstractions.jl. All
of the AD and optimization capabilities in our
tools, MadNLP.jl and SIMDiff.jl, are implemented in the Julia Language,
by leveraging its kernel and array programming capabilities.

@@

\subsection{Abstraction}
The SIMD abstraction under consideration is as follows:
\begin{align}\label{eqn:prob}
  \min_{x^\flat\leq x \leq x^\sharp}
  & \sum_{l\in[L]}\sum_{i\in [I_l]} f^{(l)}(x; p^{(l)}_i)\\\nonumber
  \st\; &\forall m\in[M]:\\\nonumber
  &\begin{aligned}[t]
    \left[g^{(m)}(x; q_j)\right]_{j\in [J_m]} +\sum_{n\in [N_m]}\sum_{k\in [K_n]}h^{(n)}(x; s^{(n)}_{k}) =0,
  \end{aligned}
  \end{align}
where $f^{(\ell)}(\cdot,\cdot)$, $g^{(m)}(\cdot,\cdot)$, and
$h^{(n)}(\cdot,\cdot)$ are twice differentiable functions with respect
to the first argument, whereas $\{\{p^{(k)}_i\}_{i\in [N_k]}\}_{k\in[K]}$,
$\{\{q^{(k)}_{i}\}_{i\in [M_l]}\}_{m\in[M]}$, and
$\{\{\{s^{(n)}_{k}\}_{k\in[K_n]}\}_{n\in[N_m]}\}_{m\in[M]}$ are
problem data, which can either be discrete or continuous.  We assume
that our functions $f^{(l)}(\cdot,\cdot)$, $g^{(m)}(\cdot,\cdot)$, and
$h^{(n)}(\cdot,\cdot)$ can be expressed with computational
graphs of moderate length. One can observe that the problem in
\eqref{eqn:prob} is expressed by the computational patterns in Section
\ref{sec:gpu}. In particular, the objective function falls within \ref{eqn:pattern-1},
the first term in the constraint falls within \ref{eqn:pattern-2}, and the second term in the constraint falls within \ref{eqn:pattern-3}.
Accordingly, the evaluation and differentiation of the
model equations in \eqref{eqn:prob} is amenable to SIMD parallelism.

@@

\subsection{Parallel AD}
Many of the physics-based models, such as AC OPF, have a highly repetitive
structure. One of the manifestations of it is that the mathematical
statement of the model is concise, even if the practical model may contain
millions of variables and constraints. This is possible due to the use of
repetition over a certain index and data sets. For example,
it suffices to use 15 computational patterns to fully specify the
AC OPF model. These patterns arise from (1) generation cost, (2) reference
bus voltage angle constraint, (3-6) active and reactive power flow (to
and from), (7) voltage angle difference constraint, (8-9) apparent
power flow limits (to and from), (10-11) power balance equations,
(12-13) generators' contributions to the power balance equations, and
(14-15) in/out flows contributions to the power balance
equations. However, such repetitive structure is not well exploited
during AD in the standard NLP modeling paradigms.  In
fact, without the SIMD abstraction, it is difficult for the AD package
to detect the parallelizable structure within the model, as it will
require the full inspection of the computational graph. By preserving
the repetitive structures in the model, one can facilitate the
implementation of parallelized derivative computations.

@@

SIMDiff.jl exploits the SIMD abstraction of the model equations to
save the computational cost spent for sparsity analysis.  This is
accomplished by applying sparsity analysis for the instruction for
each computational pattern and expand the obtained sparsity pattern
over the data over which the instruction is executed. Specifically,
the sparsity analysis code exploits Julia's multiple dispatch feature
to obtain a parameterized sparsity pattern for each instruction, and the
obtained parameterized sparsity pattern is materialized once the data
array is given. This process drastically saves in computational cost
for the sparsity analysis.
